Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2019-02-11 AnalyticStructure_main.tex
Typo removed

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can be expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Given that we employ real functions both for $p$ and $s$ is real we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{ps}^{qs'} = (H_{0 ss'}-\eps_p\delta_{ss'})\delta_{pq} + U[\decay{s}{p};\excite{q}{s'}] \;,\nn \\
& D_{ps}^{qs'} = U[\decay{s}{p};\decay{s'}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.

>>>>

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can be expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Given that we employ real functions both for $p$ and $s$ we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{ps}^{qs'} = (H_{0 ss'}-\eps_p\delta_{ss'})\delta_{pq} + U[\decay{s}{p};\excite{q}{s'}] \;,\nn \\
& D_{ps}^{qs'} = U[\decay{s}{p};\decay{s'}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.
